---
title: 'Time Series Project EDA: Fall 2022'
output: html_document
date: '2022-11-19'
---

```{r setup, include=FALSE}
library(tswge)
library(GGally)
library(tidyverse)
library(ggplot2)
library(tseries)
library(vars)
```

## Group Members:
Gowthan Katta
Josh Mitchell

## Requirements
EDA: (10 % of overall grade in the class.)
Sunday, November 19th at 11:59pm
Deliverable:
1.	4-minute YouTube (or Zoom) video:
  a.	Identify both group members. (Polished intro.) 
  b.	Describe Data Set / Time Series (Who, What, When, Where, Why and How)
  c.	Stationary / Non-Stationary 
  d.	ACFs and Spectral Densities just to explore
  e.	At least 2 candidate ARMA / ARIMA models
    a.	The models in factored form or at least separate the stationary and non-stationary factors with standard deviation or variance of the white noise.
    b.	AIC
    c.	Root ASE (short and long term forecasts)
    d.	Rolling Window RMSE (short and long term forecasts)
    e.	Visualization of Forecasts for both the short- and long-term Horizons. 
  f.	Strategy / Plans for the rest of the analysis.
  g.	This video can be completed together in one video or separately where each group member makes their own video. 
2.	Submit your slides to 2DS 
3.	Make sure your video URL is on the Google Doc. 
Google Doc Link: 
https://docs.google.com/document/d/1QbFJZJtO0Qql--08r7HF6i8gLgsFL03J552wDoAmv6U/edit?usp=sharing

## Data Description
The data set consists of stock data between 2006-01-03 and 2022-11-14. The data was collected from 29 stocks and contains 123163 observagtions. The PerChange and PerSwing variables were generated from the data and Close was chosen as the response. No NA values were observed in the data set. Descriptions of the 10 variables contained within the data set are outlined below. 

  Date - In format: yyyy-mm-dd
  High - Highest price reached in the day in USD
  Low - Lowest price reached in the day in USD
  Open - Price of the stock at market open in USD
  Close - Price of the stock at market close in USD
  Volume - Number of shares traded
  Adj.Close - Closing price adjusted for splits and dividend distributions
  Name - Stock ticker name (29 different stocks)
  PerChange - Percentage change in stock price between Open and Close. 
  PerSwing - Percentage change in stock price between High and Low. 

As shown in the pair plots below, the explanatory variables High, Low, Open, Close, and Adj.Close are all highly correlated with each other. Additionally, when investigating the pair plots, different stock tickers appear to have different distributions of the data across the explanatory variables. This appears indicate that separate models (one for each stock ticker) may be appropriate. 

```{r}
data  = read.csv(file.choose())
head(data)
tail(data)
dim(data) # 123163 observations across 8 explanatory variables
data <- drop_na(data) # no NA values 
dim(data) # 123163 observations across 8 explanatory variables
data["PerChange"] = ((data$Close - data$Open) / data$Open) * 100 # new variable
data["PerSwing"] = ((data$High - data$Low) / data$Low) * 100 # new variable
col_names = names(data)
good_names = c("High","Low","Open","Close","Volume","Adj.Close","PerChange","PerSwing")
tickers <- levels(as.factor(data$Name))

for (ticker in as.character(tickers)){
  print(data[data$Name == ticker, ] %>% ggpairs(columns=good_names, title=paste('Pair Plot for ticker: ',ticker)))
}

```

## Stationarity Check

The data for High, Low, Open, Close, Adj.Close, and Volume shows non-cyclic wandering behavior with strong correlation between data values that are near each other in time. PerChange indicates a negative autocorrelation at lag 1, and PerSwing shows slowly damping autocorrelations with possible cyclic components. PerChange appears non-stationary due to violations of the constant autocorrelations assumption. PerSwing appears non-stationary due to slowly damping autocorrelations. 

```{r}
# investigate the AAPL stock ticker
aapl <- data[data$Name == c('AAPL'),]
aapl <- aapl[ , -which(names(aapl) %in% c("Name"))]

# High, Low, Open, Close, Adj.Close, and Volume all have very similar ACF and spectral density

# High stationarity check 
# Violation of condition 1 (mean appears dependent on time) 
plotts.sample.wge(aapl$High)
acf(aapl$High, plot=T)
acf(aapl$High[1:length(aapl$High)/2], plot=T)
acf(aapl$High[((length(aapl$High)/2)+1):length(aapl$High)], plot=T)
adf.test(aapl$High) # ftr (non-stationary)

# Low stationarity check 
# Violation of condition 1 (mean appears dependent on time) 
plotts.sample.wge(aapl$Low)
acf(aapl$Low, plot=T)
acf(aapl$Low[1:length(aapl$Low)/2], plot=T)
acf(aapl$Low[((length(aapl$Low)/2)+1):length(aapl$Low)], plot=T)
adf.test(aapl$Low) # ftr (non-stationary)

# Open stationarity check 
# Violation of condition 1 (mean appears dependent on time) 
plotts.sample.wge(aapl$Open)
acf(aapl$Open, plot=T)
acf(aapl$Open[1:length(aapl$Open)/2], plot=T)
acf(aapl$Open[((length(aapl$Open)/2)+1):length(aapl$Open)], plot=T)
adf.test(aapl$Open) # ftr (non-stationary)

# Close stationarity check 
# Violation of condition 1 (mean appears dependent on time) 
plotts.sample.wge(aapl$Close)
acf(aapl$Close, plot=T)
acf(aapl$Close[1:length(aapl$Close)/2], plot=T)
acf(aapl$Close[((length(aapl$Close)/2)+1):length(aapl$Close)], plot=T)
adf.test(aapl$Close) # ftr (non-stationary)

# Adj.Close stationarity check 
# Violation of condition 1 (mean appears dependent on time) 
plotts.sample.wge(aapl$Adj.Close)
acf(aapl$Adj.Close, plot=T)
acf(aapl$Adj.Close[1:length(aapl$Adj.Close)/2], plot=T)
acf(aapl$Adj.Close[((length(aapl$Adj.Close)/2)+1):length(aapl$Adj.Close)], plot=T)
adf.test(aapl$Adj.Close)# ftr (non-stationary)

# Volume stationarity check 
# Violation of condition 1 (mean appears dependent on time) 
plotts.sample.wge(aapl$Volume)
acf(aapl$Volume, plot=T)
acf(aapl$Volume[1:length(aapl$Volume)/2], plot=T)
acf(aapl$Volume[((length(aapl$Volume)/2)+1):length(aapl$Volume)], plot=T)
adf.test(aapl$Volume) # falsely rejects Ho

# percentage change stationarity check 
plotts.sample.wge(aapl$PerChange)
acf(aapl$PerChange, plot=T)
# violation of constant autocorrelations
acf(aapl$PerChange[1:length(aapl$PerChange)/2], plot=T)
acf(aapl$PerChange[((length(aapl$PerChange)/2)+1):length(aapl$PerChange)], plot=T)
adf.test(aapl$PerChange) # falsely rejects Ho

# percentage swing stationarity check 
plotts.sample.wge(aapl$PerSwing)
acf(aapl$PerSwing, plot=T)
acf(aapl$PerSwing[1:length(aapl$PerSwing)/2], plot=T)
acf(aapl$PerSwing[((length(aapl$PerSwing)/2)+1):length(aapl$PerSwing)], plot=T)
adf.test(aapl$PerSwing) # falsely rejects Ho
```

## ARIMA Model 1
Top model is ARIMA(5,1,1)


```{r}
plotts.sample.wge(aapl$Close)
slice = aapl$Close[(length(aapl$Close)-364):length(aapl$Close)]
# take out 1-b
d1 = artrans.wge(slice, phi.tr=1)
# inspect plots for stationarity
adf.test(d1) # dicky-fuller rejects null, suggesting stationary

# smallest AIC is 2.209449 with arma(6,3)
#6,3 - 2.209449	
#7,3 - 2.212574	
#5,1 - 2.213120	
#10,2 - 2.215767	
#4,5 - 2.216730	
aic5.wge(d1, p=0:12, q=0:5) 
aic5.wge(d1, p=0:12, q=0:5,type='bic')# 0,0; 0,1; 1,0; 1,1; 2,0

est.ar.wge(d1, p=12, method='burg') # over fit 
x63 = est.arma.wge(d1, p=6,q=3)
x73 = est.arma.wge(d1, p=7,q=3)
x51 = est.arma.wge(d1, p=5,q=1)
x102 = est.arma.wge(d1, p=10,q=2)
x45 = est.arma.wge(d1, p=4,q=5)
x121 =est.arma.wge(d3, p=12,q=1)


plot(x63$res, type='p') # inspect residuals for white noise
ljung.wge(x63$res, p=6,q=3) # p-values for ljung-box test indicate residuals are sufficiently whitened
ljung.wge(x63$res, p=6,q=3, K=48)

plot(x73$res, type='p') # inspect residuals for white noise
ljung.wge(x73$res, p=6,q=3) # p-values for ljung-box test indicate residuals are sufficiently whitened
ljung.wge(x73$res, p=6,q=3, K=48)

plot(x51$res, type='p') # inspect residuals for white noise
ljung.wge(x51$res, p=6,q=3) # p-values for ljung-box test indicate residuals are sufficiently whitened
ljung.wge(x51$res, p=6,q=3, K=48)

plot(x102$res, type='p') # inspect residuals for white noise
ljung.wge(x102$res, p=6,q=3) # p-values for ljung-box test indicate residuals are sufficiently whitened
ljung.wge(x102$res, p=6,q=3, K=48)

plot(x45$res, type='p') # inspect residuals for white noise
ljung.wge(x45$res, p=6,q=3) # p-values for ljung-box test indicate residuals are sufficiently whitened
ljung.wge(x45$res, p=6,q=3, K=48)

# generating data and comparing to original shows similar structure 
plotts.sample.wge(d1)
plotts.sample.wge(gen.arma.wge(n=365, phi=x51$phi, theta = x51$theta, mu=x51$xbar, vara=x51$avar, sn=2))

# forecasting Close 5 days in the future (best rmse 3.73 on arma(5,1))
roll.win.rmse.wge(slice, horizon=5, s=0, d=1,phi=x63$phi, theta = x63$theta) # rmse 6.151
roll.win.rmse.wge(slice, horizon=5, s=0, d=1,phi=x73$phi, theta = x73$theta) # rmse 5.676
roll.win.rmse.wge(slice, horizon=5, s=0, d=1,phi=x51$phi, theta = x51$theta) # rmse 4.455
roll.win.rmse.wge(slice, horizon=5, s=0, d=1,phi=x102$phi, theta = x102$theta) # rmse 5.649
roll.win.rmse.wge(slice, horizon=5, s=0, d=1,phi=x45$phi, theta = x45$theta) # rmse 4.864
fore_short = fore.arima.wge(slice, phi=x51$phi, theta = x51$theta, s=0, d=1, n.ahead=5, lastn=T, plot=T, limits=T)
fore_5day_ASE = mean((slice[361:365] - fore_short$f)^2)
fore_5day_ASE # 57.6851

# forecasting Close 30 days in the future (rmse 10.184)
roll.win.rmse.wge(slice, horizon=30, s=0, d=1,phi=x63$phi, theta = x63$theta) # rmse 10.678
roll.win.rmse.wge(slice, horizon=30, s=0, d=1,phi=x73$phi, theta = x73$theta) # rmse 10.521
roll.win.rmse.wge(slice, horizon=30, s=0, d=1,phi=x51$phi, theta = x51$theta) # rmse 10.184
roll.win.rmse.wge(slice, horizon=30, s=0, d=1,phi=x102$phi, theta = x102$theta) # rmse 10.259
roll.win.rmse.wge(slice, horizon=30, s=0, d=1,phi=x45$phi, theta = x45$theta) # rmse 10.283
fore_long = fore.arima.wge(slice, phi=x51$phi, theta = x51$theta, s=0, d=1, n.ahead=30, lastn=T, plot=T, limits=T)
fore_30_ASE = mean((slice[336:365] - fore_long$f)^2)
fore_30_ASE # 33.13971

x51
#$phi
#[1] -0.91728222 -0.08587382 -0.09547011 -0.01902451  0.08470061
#$theta
#[1] -0.9326259
#$avar
#[1] 8.799178

# forecasts for 5 and 30 days into the future
fore.arima.wge(slice, phi=x51$phi, theta = x51$theta, s=0, d=1, n.ahead=5, lastn=F, plot=T, limits=T)
fore.arima.wge(slice, phi=x51$phi, theta = x51$theta, s=0, d=1, n.ahead=30, lastn=F, plot=T, limits=T)
```

## ARIMA Model 2
Top ARIMA model is ARIMA(10,1,2)s=5
```{r}
plotts.sample.wge(aapl$Close)
slice = aapl$Close[(length(aapl$Close)-364):length(aapl$Close)]

plotts.sample.wge(slice)
# 1-b
d1 = artrans.wge(slice, phi.tr = c(1))
# 1-b^5
d2 = artrans.wge(d1, phi.tr=c(0,0,0,0,1))
plotts.sample.wge(d2)
#11	0	2.538508	
#10	1	2.539186	
#10	0	2.541134	
#10	2	2.543522	
#12	0	2.544000	
aic5.wge(d2, p=0:12, q=0:5) 
#10	0	2.660122	
#11	0	2.668312	
#10	1	2.668991	
#10	2	2.684143	
#12	0	2.684622	
aic5.wge(d2, p=0:12, q=0:5, type='bic') 

x110 = est.arma.wge(d2, p=11, q=0)
x101 = est.arma.wge(d2, p=10, q=1)
x100 = est.arma.wge(d2, p=10, q=0)
x102 = est.arma.wge(d2, p=10, q=2)
x120 = est.arma.wge(d2, p=12, q=0)

roll.win.rmse.wge(slice, horizon=5, s=5, d=1,phi=x110$phi, theta = x110$theta)# rmse 4.998
roll.win.rmse.wge(slice, horizon=5, s=5, d=1,phi=x101$phi, theta = x101$theta)# rmse 4.978
roll.win.rmse.wge(slice, horizon=5, s=5, d=1,phi=x100$phi, theta = x100$theta)# rmse 4.988
roll.win.rmse.wge(slice, horizon=5, s=5, d=1,phi=x102$phi, theta = x102$theta)# rmse 4.968
roll.win.rmse.wge(slice, horizon=5, s=5, d=1,phi=x120$phi, theta = x120$theta)# rmse 4.999

fore_short = fore.arima.wge(slice, phi=x102$phi, theta = x112$theta, s=5, d=1, n.ahead=5, lastn=T, plot=T, limits=T)
fore_5day_ASE = mean((slice[361:365] - fore_short$f)^2)
fore_5day_ASE # 91.59743

roll.win.rmse.wge(slice, horizon=30, s=5, d=1,phi=x110$phi, theta = x110$theta)# rmse 15.345
roll.win.rmse.wge(slice, horizon=30, s=5, d=1,phi=x101$phi, theta = x101$theta)# rmse 15.092
roll.win.rmse.wge(slice, horizon=30, s=5, d=1,phi=x100$phi, theta = x100$theta)# rmse 15.084
roll.win.rmse.wge(slice, horizon=30, s=5, d=1,phi=x102$phi, theta = x102$theta)# rmse 15.102
roll.win.rmse.wge(slice, horizon=30, s=5, d=1,phi=x120$phi, theta = x120$theta)# rmse 15.338
fore_long = fore.arima.wge(slice, phi=x102$phi, theta = x102$theta, s=11, d=1, n.ahead=30, lastn=T, plot=T, limits=T)
fore_30_ASE = mean((slice[336:365] - fore_long$f)^2)
fore_30_ASE # 134.1506

x102
#$phi
#-0.31335351 -0.15301447  0.01728017 -0.04708357 -0.55525924 -0.22591524 -0.03965276 -0.09632039 -0.01246354 -0.33384230
#$theta
# -0.30206753 -0.08957349
#$avar
#[1] 11.83544

# forecasts for 5 and 30 days into the future
fore.arima.wge(slice, phi=x102$phi, theta = x102$theta, s=5, d=1, n.ahead=5, lastn=F, plot=T, limits=T)
fore.arima.wge(slice, phi=x102$phi, theta = x102$theta, s=5, d=1, n.ahead=30, lastn=F, plot=T, limits=T)
```

## Strategy
The strategy going forward is to incorporate the additional variables into VAR and MLP models. 

```{r}
aapl_High = aapl$High[(length(aapl$High)-364):length(aapl$High)]
aapl_Low = aapl$Low[(length(aapl$Low)-364):length(aapl$Low)]
aapl_Close = aapl$Close[(length(aapl$Close)-364):length(aapl$Close)]
aapl_AdjClose = aapl$Adj.Close[(length(aapl$Adj.Close)-364):length(aapl$Adj.Close)]
aapl_Open = aapl$Open[(length(aapl$Open)-364):length(aapl$Open)]
aapl_Volume = aapl$Volume[(length(aapl$Volume)-364):length(aapl$Volume)]
aapl_PerChange = aapl$PerChange[(length(aapl$PerChange)-364):length(aapl$PerChange)]
aapl_PerSwing = aapl$PerSwing[(length(aapl$PerSwing)-364):length(aapl$PerSwing)]
plotts.sample.wge(aapl_High)
aapl_High.d1 = artrans.wge(aapl_High, phi.tr = 1)
aapl_Low.d1 = artrans.wge(aapl_Low, phi.tr = 1)
aapl_Open.d1 = artrans.wge(aapl_Open, phi.tr = 1)
aapl_AdjClose.d1 = artrans.wge(aapl_AdjClose, phi.tr = 1)
aapl_Close.d1 = artrans.wge(aapl_Close, phi.tr=1)
aapl_Volume.d1 = artrans.wge(aapl_Volume, phi.tr=1)
aapl_PerChange.d1 = artrans.wge(aapl_PerChange, phi.tr=1)
aapl_PerSwing.d1 = artrans.wge(aapl_PerSwing, phi.tr=1)

# varselect bic picks ar 2
VARselect(cbind(aapl_Close.d1, aapl_High.d1, aapl_Low.d1, aapl_Open.d1,aapl_AdjClose.d1, aapl_Volume.d1, aapl_PerChange.d1, aapl_PerSwing.d1), lag.max = 12, type='both')

var_sm = VAR(cbind(aapl_Close.d1, aapl_High.d1, aapl_Low.d1, aapl_Open.d1,aapl_AdjClose.d1, aapl_Volume.d1, aapl_PerChange.d1, aapl_PerSwing.d1), p=2, type='both')
preds = predict(var_sm, n.ahead=5)

begin = aapl$Close[(length(aapl$Close)-4):length(aapl$Close)]

closeFcst = preds$fcst$aapl_Close.d1[,1:3]+begin
plot(seq(1,365,1),aapl$Close[(length(aapl$Close)-364):length(aapl$Close)], type='l', xlim=c(0,365), ylab='Stock Prediction', main='5 day stock prediction')
lines(seq(361,365,1), closeFcst[,1], type='l', col='red')

plot(aapl$Close[(length(aapl$Close)-364):length(aapl$Close)], type='l')
aapl$Close[(length(aapl$Close)-4):length(aapl$Close)]

var_ase = mean((aapl$Close[(length(aapl$Close)-4):length(aapl$Close)] - closeFcst[,1])^2)
var_ase # 0.06560871

preds = predict(var_sm, n.ahead=30)

```