---
title: 'Time Series Project EDA: Fall 2022'
output:
  pdf_document: default
  html_document: default
  word_document: default
date: '2022-11-19'
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Group Members
Gowthan Katta
Josh Mitchell

## Project Information
All project related files an documentation can be found at: 
https://github.com/bx-lr/timeseries

Presentation can be found at: 
https://youtu.be/Wkc7MB5vu1U

## Data Description
The data set consists of stock data between 2006-01-03 and 2022-11-14. The data was collected from 29 stocks and contains 123163 observagtions. The PerChange and PerSwing variables were generated from the data and Close was chosen as the response. No NA values were observed in the data set. Descriptions of the 10 variables contained within the data set are outlined below. 

  Date - In format: yyyy-mm-dd
  High - Highest price reached in the day in USD
  Low - Lowest price reached in the day in USD
  Open - Price of the stock at market open in USD
  Close - Price of the stock at market close in USD
  Volume - Number of shares traded
  Adj.Close - Closing price adjusted for splits and dividend distributions
  Name - Stock ticker name (29 different stocks)
  PerChange - Percentage change in stock price between Open and Close. 
  PerSwing - Percentage change in stock price between High and Low. 

As shown in the pair plots below, the explanatory variables High, Low, Open, Close, and Adj.Close are all highly correlated with each other. Additionally, when investigating the pair plots, different stock tickers appear to have different distributions of the data across the explanatory variables. This appears indicate that separate models (one for each stock ticker) may be appropriate. 

```{r message=FALSE, warning=FALSE}
# required libraries
library(tswge)
library(GGally)
library(tidyverse)
library(ggplot2)
library(tseries)
library(vars)
library(forecast)
library(dplyr)
library(urca)
library(orcutt)
library(nnfor)
```

```{r message=FALSE, warning=FALSE}
data  = read.csv('data/all_stocks_2006-01-01_to_2022-11-14.csv')
head(data)
tail(data)
dim(data) # 123163 observations across 8 explanatory variables
data <- drop_na(data) # no NA values 
dim(data) # 123163 observations across 8 explanatory variables
data["PerChange"] = ((data$Close - data$Open) / data$Open) * 100 # new variable
data["PerSwing"] = ((data$High - data$Low) / data$Low) * 100 # new variable
col_names = names(data)
good_names = c("High","Low","Open","Close","Volume","Adj.Close","PerChange","PerSwing")
tickers <- levels(as.factor(data$Name))

for (ticker in as.character(tickers)){
  print(data[data$Name == ticker, ] %>% ggpairs(columns=good_names, title=paste('Pair Plot for ticker: ',ticker)))
}

```

## Stationarity Check

The data for High, Low, Open, Close, Adj.Close, and Volume shows non-cyclic wandering behavior with strong correlation between data values that are near each other in time. PerChange indicates a negative autocorrelation at lag 1, and PerSwing shows slowly damping autocorrelations with possible cyclic components. PerChange appears non-stationary due to violations of the constant autocorrelations assumption. PerSwing appears non-stationary due to slowly damping autocorrelations. 

```{r message=FALSE, warning=FALSE}
# investigate the AAPL stock ticker
aapl <- data[data$Name == c('AAPL'),]
aapl <- aapl[ , -which(names(aapl) %in% c("Name"))]

# High, Low, Open, Close, Adj.Close, and Volume all have very similar ACF and spectral density

# High stationarity check 
# Violation of condition 1 (mean appears dependent on time) 
plotts.sample.wge(aapl$High)
acf(aapl$High, plot=T)
acf(aapl$High[1:length(aapl$High)/2], plot=T)
acf(aapl$High[((length(aapl$High)/2)+1):length(aapl$High)], plot=T)
adf.test(aapl$High) # ftr (non-stationary)

# Low stationarity check 
# Violation of condition 1 (mean appears dependent on time) 
plotts.sample.wge(aapl$Low)
acf(aapl$Low, plot=T)
acf(aapl$Low[1:length(aapl$Low)/2], plot=T)
acf(aapl$Low[((length(aapl$Low)/2)+1):length(aapl$Low)], plot=T)
adf.test(aapl$Low) # ftr (non-stationary)

# Open stationarity check 
# Violation of condition 1 (mean appears dependent on time) 
plotts.sample.wge(aapl$Open)
acf(aapl$Open, plot=T)
acf(aapl$Open[1:length(aapl$Open)/2], plot=T)
acf(aapl$Open[((length(aapl$Open)/2)+1):length(aapl$Open)], plot=T)
adf.test(aapl$Open) # ftr (non-stationary)

# Close stationarity check 
# Violation of condition 1 (mean appears dependent on time) 
plotts.sample.wge(aapl$Close)
acf(aapl$Close, plot=T)
acf(aapl$Close[1:length(aapl$Close)/2], plot=T)
acf(aapl$Close[((length(aapl$Close)/2)+1):length(aapl$Close)], plot=T)
adf.test(aapl$Close) # ftr (non-stationary)

# Adj.Close stationarity check 
# Violation of condition 1 (mean appears dependent on time) 
plotts.sample.wge(aapl$Adj.Close)
acf(aapl$Adj.Close, plot=T)
acf(aapl$Adj.Close[1:length(aapl$Adj.Close)/2], plot=T)
acf(aapl$Adj.Close[((length(aapl$Adj.Close)/2)+1):length(aapl$Adj.Close)], plot=T)
adf.test(aapl$Adj.Close)# ftr (non-stationary)

# Volume stationarity check 
# Violation of condition 1 (mean appears dependent on time) 
plotts.sample.wge(aapl$Volume)
acf(aapl$Volume, plot=T)
acf(aapl$Volume[1:length(aapl$Volume)/2], plot=T)
acf(aapl$Volume[((length(aapl$Volume)/2)+1):length(aapl$Volume)], plot=T)
adf.test(aapl$Volume) # falsely rejects Ho

# percentage change stationarity check 
plotts.sample.wge(aapl$PerChange)
acf(aapl$PerChange, plot=T)
# violation of constant autocorrelations
acf(aapl$PerChange[1:length(aapl$PerChange)/2], plot=T)
acf(aapl$PerChange[((length(aapl$PerChange)/2)+1):length(aapl$PerChange)], plot=T)
adf.test(aapl$PerChange) # falsely rejects Ho

# percentage swing stationarity check 
plotts.sample.wge(aapl$PerSwing)
acf(aapl$PerSwing, plot=T)
acf(aapl$PerSwing[1:length(aapl$PerSwing)/2], plot=T)
acf(aapl$PerSwing[((length(aapl$PerSwing)/2)+1):length(aapl$PerSwing)], plot=T)
adf.test(aapl$PerSwing) # falsely rejects Ho
```
## Model 1: ARIMA(5,1,1) - Restricted to last 365-day
Top model is ARIMA(5,1,1). We factored out a 1-B term and performed a Dicky-Fuller test which suggested the data was stationary. Next, we selected the top models using the AIC and BIC. Each model was tested with the Ljung-Box test for whiteness of the residuals.  Afterwards, we calculated the rolling window RMSE for each model with both long and short term forecasts. The rolling window RMSE identified the ARIMA(5,1,1) model as performing the best. Inspecting the results, we generated a realization from the ARIMA(5,1,1) and compared it to the 365 day realization from the data set. Finally, we generated 5 and 30 day forecasts from the model.    

```{r message=FALSE, warning=FALSE}
# re read the data because r is dumb
data  = read.csv('data/all_stocks_2006-01-01_to_2022-11-14.csv')
head(data)
tail(data)
dim(data) # 123163 observations across 8 explanatory variables
data <- drop_na(data) # no NA values 
dim(data) # 123163 observations across 8 explanatory variables
data["PerChange"] = ((data$Close - data$Open) / data$Open) * 100 # new variable
data["PerSwing"] = ((data$High - data$Low) / data$Low) * 100 # new variable
col_names = names(data)
good_names = c("High","Low","Open","Close","Volume","Adj.Close","PerChange","PerSwing")
tickers <- levels(as.factor(data$Name))

# investigate the AAPL stock ticker
aapl <- data[data$Name == c('AAPL'),]
aapl <- aapl[ , -which(names(aapl) %in% c("Name"))]

#plotts.sample.wge(aapl$Close)
slice = aapl$Close[(length(aapl$Close)-364):length(aapl$Close)]
# take out 1-b
plotts.sample.wge(slice)
d1 = artrans.wge(slice, phi.tr=1)
acf(d1)
# inspect plots for stationarity

adf.test(d1) # dicky-fuller rejects null, suggesting stationary

# smallest AIC is 2.209449 with arma(6,3)
#6,3 - 2.209449	
#7,3 - 2.212574	
#5,1 - 2.213120	
#10,2 - 2.215767	
#4,5 - 2.216730	
aic5.wge(d1, p=0:12, q=0:5) 
aic5.wge(d1, p=0:12, q=0:5,type='bic')# 0,0; 0,1; 1,0; 1,1; 2,0

est.ar.wge(d1, p=12, method='burg') # over fit 
x63 = est.arma.wge(d1, p=6,q=3)
x73 = est.arma.wge(d1, p=7,q=3)
x51 = est.arma.wge(d1, p=5,q=1)
x102 = est.arma.wge(d1, p=10,q=2)
x45 = est.arma.wge(d1, p=4,q=5)
x121 =est.arma.wge(d1, p=12,q=1)


plot(x63$res, type='p') # inspect residuals for white noise
ljung.wge(x63$res, p=6,q=3) # p-values for ljung-box test indicate residuals are sufficiently whitened
ljung.wge(x63$res, p=6,q=3, K=48)

plot(x73$res, type='p') # inspect residuals for white noise
ljung.wge(x73$res, p=6,q=3) # p-values for ljung-box test indicate residuals are sufficiently whitened
ljung.wge(x73$res, p=6,q=3, K=48)

plot(x51$res, type='p') # inspect residuals for white noise
ljung.wge(x51$res, p=6,q=3) # p-values for ljung-box test indicate residuals are sufficiently whitened
ljung.wge(x51$res, p=6,q=3, K=48)

plot(x102$res, type='p') # inspect residuals for white noise
ljung.wge(x102$res, p=6,q=3) # p-values for ljung-box test indicate residuals are sufficiently whitened
ljung.wge(x102$res, p=6,q=3, K=48)

plot(x45$res, type='p') # inspect residuals for white noise
ljung.wge(x45$res, p=6,q=3) # p-values for ljung-box test indicate residuals are sufficiently whitened
ljung.wge(x45$res, p=6,q=3, K=48)


# forecasting Close 5 days in the future (best rmse 3.73 on arma(5,1))
roll.win.rmse.wge(slice, horizon=5, s=0, d=1,phi=x63$phi, theta = x63$theta) # rmse 6.151
roll.win.rmse.wge(slice, horizon=5, s=0, d=1,phi=x73$phi, theta = x73$theta) # rmse 5.676
roll.win.rmse.wge(slice, horizon=5, s=0, d=1,phi=x51$phi, theta = x51$theta) # rmse 4.455
roll.win.rmse.wge(slice, horizon=5, s=0, d=1,phi=x102$phi, theta = x102$theta) # rmse 5.649
roll.win.rmse.wge(slice, horizon=5, s=0, d=1,phi=x45$phi, theta = x45$theta) # rmse 4.864
fore_short = fore.arima.wge(slice, phi=x51$phi, theta = x51$theta, s=0, d=1, n.ahead=5, lastn=T, plot=T, limits=T)
fore_5day_ASE = mean((slice[361:365] - fore_short$f)^2)
fore_5day_ASE # 57.6851

# forecasting Close 30 days in the future (rmse 10.184)
roll.win.rmse.wge(slice, horizon=30, s=0, d=1,phi=x63$phi, theta = x63$theta) # rmse 10.678
roll.win.rmse.wge(slice, horizon=30, s=0, d=1,phi=x73$phi, theta = x73$theta) # rmse 10.521
roll.win.rmse.wge(slice, horizon=30, s=0, d=1,phi=x51$phi, theta = x51$theta) # rmse 10.184
roll.win.rmse.wge(slice, horizon=30, s=0, d=1,phi=x102$phi, theta = x102$theta) # rmse 10.259
roll.win.rmse.wge(slice, horizon=30, s=0, d=1,phi=x45$phi, theta = x45$theta) # rmse 10.283
fore_long = fore.arima.wge(slice, phi=x51$phi, theta = x51$theta, s=0, d=1, n.ahead=30, lastn=T, plot=T, limits=T)
fore_30_ASE = mean((slice[336:365] - fore_long$f)^2)
fore_30_ASE # 33.13971

# generating data and comparing to original shows similar structure 
plotts.sample.wge(d1)
plotts.sample.wge(gen.arma.wge(n=365, phi=x51$phi, theta = x51$theta, mu=x51$xbar, vara=x51$avar, sn=2))

x51
#$phi
#[1] -0.91728222 -0.08587382 -0.09547011 -0.01902451  0.08470061
#$theta
#[1] -0.9326259
#$avar
#[1] 8.799178

# forecasts for 5 and 30 days into the future
fore.arima.wge(slice, phi=x51$phi, theta = x51$theta, s=0, d=1, n.ahead=5, lastn=F, plot=T, limits=T)
fore.arima.wge(slice, phi=x51$phi, theta = x51$theta, s=0, d=1, n.ahead=30, lastn=F, plot=T, limits=T)
```
## Model 2: ARMA(10,5) - Utilizing All Days
For our Model 2, we wanted to perform an evaluation on the entire dataset. First, we performed an OLS fit on the Close and the Time. We found that the null hypothesis was rejected and that the time was found to be significant. Second, we used the previous OLS fit on a Cochrane-Orcutt test. We found that we failed to reject the null hypothesis, which meant that time did NOT appear to be significant. As such, there is a small chance (or evidence) of a deterministic trend in the data.

Next, we evaluated the top AIC value and came upon our AR(10)MA(5) model. We used the parameters from the model and fit on the last 365 days (1 year) and found that our values were slightly smaller compared to the previous model in which we included a seasonal and deterministic value. 

```{r message=FALSE, warning=FALSE}

# re read the data because r is dumb
data  = read.csv('data/all_stocks_2006-01-01_to_2022-11-14.csv')
head(data)
tail(data)
dim(data) # 123163 observations across 8 explanatory variables
data <- drop_na(data) # no NA values 
dim(data) # 123163 observations across 8 explanatory variables
data["PerChange"] = ((data$Close - data$Open) / data$Open) * 100 # new variable
data["PerSwing"] = ((data$High - data$Low) / data$Low) * 100 # new variable
col_names = names(data)
good_names = c("High","Low","Open","Close","Volume","Adj.Close","PerChange","PerSwing")
tickers <- levels(as.factor(data$Name))

# investigate the AAPL stock ticker
aapl <- data[data$Name == c('AAPL'),]
aapl <- aapl[ , -which(names(aapl) %in% c("Name"))]
# convert the data to timeseries object
aapl.ts <- as.ts(aapl$Close)

# visualize the realization
plotts.wge(aapl.ts)
acf(aapl.ts)

# look at aic, bic, and bicc models
aic5.wge(aapl.ts)
aic5.wge(aapl.ts, type = 'bic')
aic5.wge(aapl.ts, type = 'aicc')

aapl$index <- 1:nrow(aapl)

# OLS: time appears to be significant (rejected the null)
fit = lm(Close~index, data = aapl)
summary(fit)

# Cochrane-Orcutt: time does NOT appear to be significant (failed to reject the null)
# Small chance (or evidence of a deterministic trend in the data)
cfit = cochrane.orcutt(fit)
summary(cfit)


aic.wge(aapl.ts, p = 0:10, q = 0:5)
aic5.wge(aapl.ts, p = 0:10, q = 0:5)



aapl.ts2 = aapl.ts[(length(aapl.ts)-364):length(aapl.ts)]
#aapl.ts
#aapl.ts2


aapl.pred1 = fore.arma.wge(aapl.ts, phi = c(-0.322309522, 1.057487180, -0.055477750, -0.844429717, 0.488404731, 0.610026678, -0.002045086, -0.002878810, 0.046165984, 0.024858653), 
                           theta = c(-1.2731792, -0.1666368, -0.2501280, -1.1441992, -0.6634130), n.ahead = 5, lastn = T, limits = F)

aapl.pred1.1 <- fore.arma.wge(aapl.ts, phi = c(-0.322309522, 1.057487180, -0.055477750, -0.844429717, 0.488404731, 0.610026678, -0.002045086, -0.002878810, 0.046165984, 0.024858653), 
                           theta = c(-1.2731792, -0.1666368, -0.2501280, -1.1441992, -0.6634130), n.ahead = 5, lastn = F, limits = T)


ASE1 = mean((aapl.pred1$f - aapl.ts2[361:365])^2)
ASE1 #53.289


aapl.pred2 = fore.arma.wge(aapl.ts, phi = c(-0.322309522, 1.057487180, -0.055477750, -0.844429717, 0.488404731, 0.610026678, -0.002045086, -0.002878810, 0.046165984, 0.024858653), 
                           theta = c(-1.2731792, -0.1666368, -0.2501280, -1.1441992, -0.6634130), n.ahead = 30, lastn = T, limits = F)

aapl.pred2.1 = fore.arma.wge(aapl.ts, phi = c(-0.322309522, 1.057487180, -0.055477750, -0.844429717, 0.488404731, 0.610026678, -0.002045086, -0.002878810, 0.046165984, 0.024858653), 
                           theta = c(-1.2731792, -0.1666368, -0.2501280, -1.1441992, -0.6634130), n.ahead = 30, lastn = F, limits = T)

ASE2 = mean((aapl.pred2$f - aapl.ts[336:365])^2)
ASE2 #71.685


#1.079
roll.win.rmse.wge(aapl.ts, horizon = 5, d=0, phi = c(-0.322309522, 1.057487180, -0.055477750, -0.844429717, 0.488404731, 0.610026678, -0.002045086, -0.002878810, 0.046165984, 0.024858653), 
                  theta = c(-1.2731792, -0.1666368, -0.2501280, -1.1441992, -0.6634130))

#2.605
roll.win.rmse.wge(aapl.ts, horizon = 30, d=0, phi = c(-0.322309522, 1.057487180, -0.055477750, -0.844429717, 0.488404731, 0.610026678, -0.002045086, -0.002878810, 0.046165984, 0.024858653), 
                  theta = c(-1.2731792, -0.1666368, -0.2501280, -1.1441992, -0.6634130))



aapl.ts2 = aapl.ts[(length(aapl.ts)-364):length(aapl.ts)]
aapl.ts2 = aapl.ts[(length(aapl.ts)-364):length(aapl.ts)]


aapl.pred1 = fore.arma.wge(aapl.ts2, phi = c(-0.322309522, 1.057487180, -0.055477750, -0.844429717, 0.488404731, 0.610026678, -0.002045086, -0.002878810, 0.046165984, 0.024858653), 
                           theta = c(-1.2731792, -0.1666368, -0.2501280, -1.1441992, -0.6634130), n.ahead = 5, lastn = T, limits = F)

aapl.pred1.1 <- fore.arma.wge(aapl.ts2, phi = c(-0.322309522, 1.057487180, -0.055477750, -0.844429717, 0.488404731, 0.610026678, -0.002045086, -0.002878810, 0.046165984, 0.024858653), 
                           theta = c(-1.2731792, -0.1666368, -0.2501280, -1.1441992, -0.6634130), n.ahead = 5, lastn = F, limits = T)


ASE1 = mean((aapl.pred1$f - aapl.ts2[361:365])^2)
ASE1 #53.289


aapl.pred2 = fore.arma.wge(aapl.ts2, phi = c(-0.322309522, 1.057487180, -0.055477750, -0.844429717, 0.488404731, 0.610026678, -0.002045086, -0.002878810, 0.046165984, 0.024858653), 
                           theta = c(-1.2731792, -0.1666368, -0.2501280, -1.1441992, -0.6634130), n.ahead = 30, lastn = T, limits = F)

aapl.pred2.1 = fore.arma.wge(aapl.ts2, phi = c(-0.322309522, 1.057487180, -0.055477750, -0.844429717, 0.488404731, 0.610026678, -0.002045086, -0.002878810, 0.046165984, 0.024858653), 
                           theta = c(-1.2731792, -0.1666368, -0.2501280, -1.1441992, -0.6634130), n.ahead = 30, lastn = F, limits = T)

ASE2 = mean((aapl.pred2$f - aapl.ts2[336:365])^2)
ASE2 #29.171


#4.404
roll.win.rmse.wge(aapl.ts2, horizon = 5, d=0, phi = c(-0.322309522, 1.057487180, -0.055477750, -0.844429717, 0.488404731, 0.610026678, -0.002045086, -0.002878810, 0.046165984, 0.024858653), 
                  theta = c(-1.2731792, -0.1666368, -0.2501280, -1.1441992, -0.6634130))

#10.142
roll.win.rmse.wge(aapl.ts2, horizon = 30, d=0, phi = c(-0.322309522, 1.057487180, -0.055477750, -0.844429717, 0.488404731, 0.610026678, -0.002045086, -0.002878810, 0.046165984, 0.024858653), 
                  theta = c(-1.2731792, -0.1666368, -0.2501280, -1.1441992, -0.6634130))

```


### VAR Model
For our VAR model, we evaluated the full data set for aapl stock with near and far-term horizons. The optimal model for near-term predictions incorporated first-difference data on all of the explanatory variables. When this model was evaluated for far-term prediction accuracy, we identified significant performance degradation. Unmodified data was then used for far-term model creation and forecasts. The Unmodified data model was observed performing better in the long-term and the differenced data was observed performing better in the short-term. 

```{r}
# difference some of the data
aapl_short_High.d1 = artrans.wge(aapl$High[1:4242], phi.tr = 1)
aapl_short_Low.d1 = artrans.wge(aapl$Low[1:4242], phi.tr = 1)
aapl_short_Open.d1 = artrans.wge(aapl$Open[1:4242], phi.tr = 1)
aapl_short_AdjClose.d1 = artrans.wge(aapl$Adj.Close[1:4242], phi.tr = 1)
aapl_short_Volume.d1 = artrans.wge(aapl$Volume[1:4242], phi.tr=1)
aapl_short_PerChange.d1 = artrans.wge(aapl$PerChange[1:4242], phi.tr=1)
aapl_short_PerSwing.d1 = artrans.wge(aapl$PerSwing[1:4242], phi.tr=1)

#
#5 day difference all but close
#

# varselect bic chooses p=5
VARselect(cbind(aapl$Close[2:4242], aapl_short_Open.d1, aapl_short_High.d1, aapl_short_Low.d1, aapl_short_AdjClose.d1, aapl_short_Volume.d1, aapl_short_PerChange.d1, aapl_short_PerSwing.d1), lag.max=30, season=30, type='both' )

# var with p=5
aapl_5 = VAR(cbind(aapl$Close[2:4242], aapl_short_Open.d1, aapl_short_High.d1, aapl_short_Low.d1, aapl_short_AdjClose.d1, aapl_short_Volume.d1, aapl_short_PerChange.d1, aapl_short_PerSwing.d1), lag.max=30, season=30, type='both', p=5 )

# get 5 day forecast
preds_5day=predict(aapl_5,n.ahead=5)
preds_5day

plot(seq(1,360,1), aapl$Close[3883:4242], type = "l",xlim = c(0,365), ylab = "Close Price", main = "5 Day Forecast", xlab='Last Year in Days') + lines(seq(361,365,1), preds_5day$fcst$X[,1], type = "l", col = "red") + lines(seq(361,365,1), aapl$Close[4243:4247], type = 'l', col='black') + lines(seq(361,365,1), preds_5day$fcst$X[ ,3], lty = 'dashed', col = 'blue') + lines(seq(361,365,1), preds_5day$fcst$X[ ,2], lty = 'dashed', col = 'blue')

ASE_5day = mean((aapl$Close[4243:4247] - preds_5day$fcst$X[,1])^2)
ASE_5day # 47.076

#
# 30 day lag all but Close
#

# 30 day forecast
aapl_long_High.d1 = artrans.wge(aapl$High[1:4217], phi.tr = 1)
aapl_long_Low.d1 = artrans.wge(aapl$Low[1:4217], phi.tr = 1)
aapl_long_Open.d1 = artrans.wge(aapl$Open[1:4217], phi.tr = 1)
aapl_long_AdjClose.d1 = artrans.wge(aapl$Adj.Close[1:4217], phi.tr = 1)
aapl_long_Volume.d1 = artrans.wge(aapl$Volume[1:4217], phi.tr=1)
aapl_long_PerChange.d1 = artrans.wge(aapl$PerChange[1:4217], phi.tr=1)
aapl_long_PerSwing.d1 = artrans.wge(aapl$PerSwing[1:4217], phi.tr=1)

# varselect bic chooses p=17 
VARselect(cbind(aapl$Close[2:4242], aapl_long_High.d1, aapl_long_Low.d1, aapl_long_Open.d1, aapl_long_AdjClose.d1, aapl_long_Volume.d1, aapl_long_PerChange.d1, aapl_long_PerSwing.d1), lag.max=30, season=30, type='both')

aapl_30 = VAR(cbind(aapl$Close[2:4242], aapl_long_High.d1, aapl_long_Low.d1, aapl_long_Open.d1, aapl_long_AdjClose.d1, aapl_long_Volume.d1, aapl_long_PerChange.d1, aapl_long_PerSwing.d1), lag.max=30, season=30, type='both', p=17)

# get 30 day predictions
preds_30day=predict(aapl_30,n.ahead=30)
preds_30day

# plot predictions with original data
plot(seq(1,335,1), aapl$Close[3883:4217], type = "l",xlim = c(0,365), ylim=c(10,200), ylab = "Close Price", main = "30 Day Forecast", xlab='Last Year in Days') + lines(seq(336,365,1), preds_30day$fcst$X[,1], type = "l", col = "red") + lines(seq(336,365,1), aapl$Close[4218:4247], type='l', col='blue')

# calculate ase
ASE_30day = mean((aapl$Close[4218:4247] - preds_30day$fcst$X[,1])^2)
ASE_30day # 3171.503

### no difference
#
# 5 day no difference
#
VARselect(cbind(aapl$Close[1:4242], aapl$Open[1:4242], aapl$High[1:4242], aapl$Low[1:4242], aapl$Adj.Close[1:4242], aapl$Volume[1:4242], aapl$PerChange[1:4242], aapl$PerSwing[1:4242]), lag.max=30, season=30, type='both' )

# var with p=3
aapl_nl_5 = VAR(cbind(aapl$Close[1:4242], aapl$Open[1:4242], aapl$High[1:4242], aapl$Low[1:4242], aapl$Adj.Close[1:4242], aapl$Volume[1:4242], aapl$PerChange[1:4242], aapl$PerSwing[1:4242]), lag.max=30, season=30, type='both', p=3 )

preds_5day=predict(aapl_nl_5,n.ahead=5)

plot(seq(1,360,1), aapl$Close[3883:4242], type = "l",xlim = c(0,365), ylab = "Close Price", main = "5 Day Forecast", xlab='Last Year in Days') + lines(seq(361,365,1), preds_5day$fcst$y1[,1], type = "l", col = "red") + lines(seq(361,365,1), aapl$Close[4243:4247], type = 'l', col='blue') + lines(seq(361,365,1), preds_5day$fcst$y1[ ,3], lty = 'dashed', col = 'blue') + lines(seq(361,365,1), preds_5day$fcst$y1[ ,2], lty = 'dashed', col = 'blue')

ASE_5day = mean((aapl$Close[4243:4247] - preds_5day$fcst$y1[,1])^2)
ASE_5day # 64.08639


#
# 30 day no difference
#
# varselect bic chooses p=2
VARselect(cbind(aapl$Close[1:4217], aapl$Open[1:4217], aapl$High[1:4217], aapl$Low[1:4217], aapl$Adj.Close[1:4217], aapl$Volume[1:4217], aapl$PerChange[1:4217], aapl$PerSwing[1:4217]), lag.max=30, season=30, type='both' )

# var with p=2
aapl_nl_30 = VAR(cbind(aapl$Close[1:4217], aapl$Open[1:4217], aapl$High[1:4217], aapl$Low[1:4217], aapl$Adj.Close[1:4217], aapl$Volume[1:4217], aapl$PerChange[1:4217], aapl$PerSwing[1:4217]), lag.max=30, season=30, type='both', p=2 )

# get 30 day predictions
preds_30day=predict(aapl_nl_30,n.ahead=30)
preds_30day
# plot predictions with original data
plot(seq(1,335,1), aapl$Close[3883:4217], type = "l",xlim = c(0,365), ylab = "Close Price", main = "30 Day Forecast", xlab='Last Year in Days') + lines(seq(336,365,1), preds_30day$fcst$y1[,1], type = "l", col = "red") + lines(seq(336,365,1), aapl$Close[4218:4247], type='l', col='black') + lines(seq(336,365,1), preds_30day$fcst$y1[ ,3], lty = 'dashed', col = 'blue') + lines(seq(336,365,1), preds_30day$fcst$y1[ ,2], lty = 'dashed', col = 'blue')

# calculate ase
ASE_30day = mean((aapl$Close[4218:4247] - preds_30day$fcst$y1[,1])^2)
ASE_30day # 30.51524
```

## MLP Model
The MLP Model was trained on the last 365 days of the Apple Stocks. We used the past 360 days and 335 days respectively in order to create the 5 and 30 day forecasts. We did this since we did not want the existing last 5 and 30 day values to contribute to the model, so we separated our dataset accordingly. First, we the “Open” variable as our explanatory variables. For the 5 day forecast, the model created a 1 ,2, and 4 day lag of this variable. For the 30 day forecast, the model created a 1and 4 day  lag of this variable. 

Second, we used the information of the forecast from the first step and implemented it into our final MLP Model. Again, we separated the 5 and 30 days from the dataset in order to not bias the forecasts. Lastly, we compared the forecasts of the final model to the actuals.

```{r}
aapl_365 <- aapl[3883:4247,]
aapl_365

aapl365_small = aapl_365[1:360,]
aapl365_smallDF = data.frame(Close = ts(aapl365_small$Close))

# Using Open
fit.mlp.1 = mlp(ts(aapl365_small$Open),reps = 50,comb = "median",xreg = aapl365_smallDF)
fit.mlp.1
plot(fit.mlp.1)


# Forecast 5
aapl365_DF = data.frame(Close = ts(aapl_365$Close))
fore.mlp.5 = forecast(fit.mlp.1, h = 5, xreg = aapl365_DF)
plot(fore.mlp.5)

plot(aapl365_small$Close, type = "l", ylab = "Close Price", main = "5 Day Forecast", xlab='Last Year in Days')
lines(seq(361,365,1),fore.mlp.5$mean,col = "blue")

ASE = mean((aapl_365$Close[361:365] - fore.mlp.5$mean)^2)
ASE


# Forecast 30
aapl365_small.2 = aapl_365[1:335,]
aapl365_smallDF.2 = data.frame(Close = ts(aapl365_small.2$Close))

fit.mlp.2 = mlp(ts(aapl365_small.2$Open),reps = 50,comb = "median",xreg = aapl365_smallDF.2)
fit.mlp.2
plot(fit.mlp.2)

fore.mlp.30 = forecast(fit.mlp.2, h = 30, xreg = aapl365_DF)
plot(fore.mlp.30)

plot(aapl365_small$Close, type = "l", ylab = "Close Price", main = "30 Day Forecast", xlab='Last Year in Days')
lines(seq(336,365,1),fore.mlp.30$mean,col = "blue")

ASE = mean((aapl_365$Close[336:365] - fore.mlp.30$mean)^2)
ASE

##### MLP Part 2#####


# Using the (Forecast) Close

fit.mlp.AC = mlp(ts(aapl365_small$Close),reps = 50, comb = "mean")
fore.mlp.AC = forecast(fit.mlp.AC, h = 5)

aapl365_smallDF_fore = data.frame(Close = ts(fore.mlp.AC$mean))
aapl365_smallDF_fore

fit.mlp = mlp(ts(aapl365_small$Open),reps = 50,comb = "mean",xreg = aapl365_DF)
fit.mlp
plot(fit.mlp)

aapl_DF = data.frame(Close = ts(c(aapl365_smallDF$Close,aapl365_smallDF_fore$Close)))


# Forecast 5

fore.mlp = forecast(fit.mlp, h = 5, xreg = aapl_DF)
plot(fore.mlp)

plot(aapl_365$Close, type = "l", ylab = "Close Price", main = "5 Day Forecast", xlab='Last Year in Days')
lines(seq(361,365,1),fore.mlp$mean,col = "blue")

ASE = mean((aapl_365$Close[361:365] - fore.mlp$mean)^2)
ASE

# Forecast 30

fore.mlp.AC.2 = forecast(fit.mlp.AC, h = 30)

aapl365_smallDF_fore.2 = data.frame(Close = ts(fore.mlp.AC.2$mean))
aapl_DF.2 = data.frame(Close = ts(c(aapl365_smallDF$Close,aapl365_smallDF_fore.2$Close)))

fore.mlp.2 = forecast(fit.mlp, h = 30, xreg = aapl_DF.2)
plot(fore.mlp.2)

plot(aapl_365$Close, type = "l", ylab = "Close Price", main = "30 Day Forecast", xlab='Last Year in Days')
lines(seq(336,365,1),fore.mlp.2$mean,col = "blue")

ASE = mean((aapl_365$Close[336:365] - fore.mlp.2$mean)^2)
ASE

```


## Ensemble
For the ensemble, we combined the forecasts from the VAR and MLP Models. Note that the forecasts from the VAR Model were created by using all of the predictors and the forecasts from the MLP Model were created using the “Open” variable and the respective lags. The Ensemble Model took the combinations of both of the above model and averaged out the predictions.

```{r}
# Ensemble 5 vs 30 Day Forecasts


ensemble = (fore.mlp$mean + preds_5day$fcst$y1[,1])/2

plot(aapl_365$Close, type = "l", ylab = "Close Price", main = "5 Day Forecast", xlab='Last Year in Days')
lines(seq(361,365,1),ensemble,col = "red")

ASE = mean((aapl_365$Close[361:365] - ensemble)^2)
ASE



ensemble2 = (fore.mlp.2$mean + preds_30day$fcst$y1[,1])/2

plot(aapl_365$Close, type = "l", ylab = "Close Price", main = "30 Day Forecast", xlab='Last Year in Days')
lines(seq(336,365,1),ensemble2,col = "red")

ASE = mean((aapl_365$Close[336:365] - ensemble2)^2)
ASE


```
## Summary
After comparing the models, we found that the MLP performed the best for the 5-Day Forecasts while the Ensemble performed the best for the 30-Day Forecasts. Although each model performed well, with different data these results could vary.

Links:
- https://youtu.be/iou3k6YS61c
- https://github.com/bx-lr/timeseries

